---
title:  "Adjusted Home Run Frequencies" 
author: Jason Osborne
output:
  pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE,eval=TRUE,message=FALSE,cache=TRUE,fig.align="center",out.width="50%")
```

# Adjusted hr frequencies, 2022.

Let us see how the batter-pitcher combination (`bpcombo`) frequencies have
varied over time

```{r plot1, out.width="50%", fig.align="center"}
load("./allyrs.12vars.RData")
allyrs.12vars %>% select(BAT_HAND_CD,PIT_HAND_CD) %>% table %>% prop.table -> bpfrqs.era
# all four combo frqs
bpfrqs.era
# batter hand
bpfrqs.era %>% rowSums()
# pitcher hand
bpfrqs.era %>% colSums()
# conditionally on batter hand, P(LHP|LHB), P(LHP|RHB)
bpfrqs.era[1,1]/sum(bpfrqs.era[1,])
bpfrqs.era[2,1]/sum(bpfrqs.era[2,])
# conditionally on pitcher hand, P(LHB|LHP), P(LHB|RHP)
bpfrqs.era[1,1]/sum(bpfrqs.era[,1])
bpfrqs.era[1,2]/sum(bpfrqs.era[,2])
```

\newpage

These four bpcombo frequencies have changed little over time, though the preference
for LHB when facing RHP may have decreased slightly.
```{r plot2, echo=FALSE}
allyrs.12vars %>% select(BAT_HAND_CD,PIT_HAND_CD,season) %>% table -> bpfrq.array

bpfrq.array %>% prop.table(margin=c("season")) -> bpfrq.byyear
plot(names(bpfrq.byyear[2,2,]),bpfrq.byyear[2,2,],ylim=c(0,0.7),type="b",lty=1,
     xlab="season",ylab="BPCombo Frequency",main="Batter/pitcher combo frqs over time",lwd=1.5)
lines(names(bpfrq.byyear[1,2,]),bpfrq.byyear[1,2,],type="b",lty=2)
lines(names(bpfrq.byyear[2,1,]),bpfrq.byyear[2,1,],type="b",lty=1,col="red")
lines(names(bpfrq.byyear[1,1,]),bpfrq.byyear[1,1,],type="b",lty=2,col="red")
legend(2012,0.7,c("RHBRHP","LHBRHP","RHBLHP","LHBLHP"),lty=c(1,2,1,2),col=c("black","black","red","red"),lwd=1.5)

allyrs.12vars %>% group_by(BAT_HAND_CD,PIT_HAND_CD) %>% summarize(count=n()) %>% 
  mutate(relfreq=count/sum(count)) -> bpfrq.vec
```
The effects of bpcombo on home run frequency can be investigated with an interaction plot, generated with data from 2012-2021:
```{r bpfrqs.era}
allyrs.12vars %>% group_by(BAT_HAND_CD,PIT_HAND_CD) %>% summarize(hrfrq=mean(hr)) -> hrfrq.bp.era
hrfrq.bp.era
ggplot(hrfrq.bp.era,aes(y=hrfrq,x=BAT_HAND_CD,color=PIT_HAND_CD)) + 
  geom_line(aes(group=PIT_HAND_CD)) + geom_point() + 
  ggtitle("BP Combo effect on P(HR), 2012-2021")
```
Looking over this 10 year period, it can be seen that home runs are least likely when
a LHB is facing a LHP.  Remarkably, the effect of the batter hand only appears 
to matter when facing lefties.  Wow!

For a given park, the frequencies of the four combinations can vary dramatically 
from one season to the next, depending upon the personnel of the home team and with
the unbalanced schedules of years past, upon the personnel of other teams in the
division.  In light of bpcombo effects, home run frequencies for each park can
be adjusted to league-wide bpcombo frequencies simply by reweighting the
four conditional home run rates to the these frequencies.  

```{r adjplot1}
allyrs.12vars %>% group_by(park,BAT_HAND_CD,PIT_HAND_CD) %>% 
  summarize(hrfrq=mean(hr)) %>% pivot_wider(values_from=hrfrq,
                                names_from=c("BAT_HAND_CD","PIT_HAND_CD")) ->
  hrsummary.wide
hrsummary.wide %>% mutate(adjhr=bpfrqs.era[1,1]*L_L + bpfrqs.era[1,2]*L_R + 
                                bpfrqs.era[2,1]*R_L + bpfrqs.era[2,2]*R_R) ->
  hrsummary.wide 
# unadjusted
allyrs.12vars %>% group_by(park) %>% summarize(obshr=mean(hr)) -> hrfrqs.bypark
hrsummary.wide %>% inner_join(hrfrqs.bypark) -> hrsummary.wide
```
These adjusted frequencies can be plotted against park, along with the unadjusted
frequencies.  Further investigation of changes over time is warranted though,
as a glance at 10-year averages still shows considerable variability in 
bpcombo frequencies across parks.  It must be kept in mind that many players
reside with the same team for long periods of time, so these 10 years are not
at all independent.  However, we average anyway ...

A technique worth mentioning in the construction of this plot is to achieve
an ordering of parks on the horizontal axis according to either the observed
or adjusted home run rate by so ordering the levels of park as a factor.
```{r orderpark}
hrsummary.wide$park <- factor(hrsummary.wide$park,
                              levels=hrsummary.wide$park[order(hrsummary.wide$adjhr)])
hrsummary.wide %>% pivot_longer(cols=c("adjhr","obshr"),values_to="phr") ->
  hrsummary.tall
```
Now `ggplot` can be used ...
```{r parkplot}
ggplot(hrsummary.tall) + geom_point(aes(y=phr,x=park,color=name)) + 
  theme(axis.text.x=element_text(angle=50)) + 
  ggtitle("HR Freqs, adj and observed \n 2012-2021")
```
